---
title: "EDA_Candice"
author: "Candice Yu"
date: "2023-11-16"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(data.table)
library(survival)
```

# Data Cleaning 
```{r}
# Define a function to convert multiple columns to factors
convert_to_factor <- function(df, columns) {
  df[columns] <- lapply(df[columns], factor)
  return(df)
}

# Load the data and preprocess it
mortality_data <- read_csv("mortality_data.csv") %>%
  janitor::clean_names() %>%
  drop_na(outcome) %>%
  convert_to_factor(., c("group", "gendera", "outcome", "hypertensive", 
                         "atrialfibrillation", "chd_with_no_mi", "diabetes", 
                         "deficiencyanemias", "depression", "hyperlipemia", 
                         "renal_failure", "copd")) %>%
  rename(gender = gendera)  # Rename gendera to gender after conversion


# Imputing numerical columns with mean value 
numerical_columns <- sapply(mortality_data, is.numeric)
mortality_data[numerical_columns] <- lapply(mortality_data[numerical_columns], function(x) {
   ifelse(is.na(x), mean(x, na.rm = TRUE), x)
})

write_csv(mortality_data, "mortality_data_cleaned.csv")
```
# Overview of the data 
```{r}
skimr::skim(mortality_data)
```
Target Variable - Outcome
* 0 - Alive
* 1 - Death

# Change the factor variables for EDA 
```{r}
# Assuming 'mortality_data' is already loaded and cleaned up as per your previous steps

# Manually recoding factors to their meaningful character values
mortality_data_EDA <- mortality_data %>%
  mutate(
    group = recode(group, `1` = "Group 1", `2` = "Group 2"),
    gender = recode(gender, `1` = "Male", `2` = "Female"),
    outcome = recode(outcome, `0` = "Alive", `1` = "Death")
  ) %>%
  mutate_if(is.factor, as.character) # Converts all remaining factors to characters

# Function to recode factor variables to 'Yes' or 'No'
convert_factors_to_yes_no <- function(df, comorbidity_columns) {
  df <- df %>%
    mutate(across(all_of(comorbidity_columns), ~ ifelse(as.character(.) == "1", "Yes", "No")))
  return(df)
}

# List of comorbidity columns to convert
comorbidity_columns <- c("hypertensive", "atrialfibrillation", "chd_with_no_mi", 
                         "diabetes", "deficiencyanemias", "depression", 
                         "hyperlipemia", "renal_failure", "copd")

# Apply the function to the mortality_data_EDA dataframe
mortality_data_EDA <- convert_factors_to_yes_no(mortality_data_EDA, comorbidity_columns)

```


# Gender & Outcome
```{r}
mortality_data_EDA %>%
  group_by(gender) %>%
  summarise(
    Count = n(), # total number of entries for each gender
    Alive = sum(outcome == "Alive"), # number of outcomes with value 0
    Death = sum(outcome == "Death"), # number of outcomes with value 1
    Percentage = Death/Count
  ) %>%
  knitr::kable(digits = 3)  
```


# Age & Outcome
```{r}
# Define age intervals
age_breaks <- c(-Inf, 20, 40, 60, 80, Inf)
age_labels <- c('Under 20', '20-40', '40-60', '60-80', 'Over 80')

# Create age groups and summarize outcomes
mortality_data_EDA %>%
  mutate(Age_group = cut(age, breaks = age_breaks, labels = age_labels, right = FALSE)) %>%
  group_by(Age_group) %>%
  summarise(Count = n(),
            Alive = sum(outcome == "Alive", na.rm = TRUE),
            Death = sum(outcome == "Death", na.rm = TRUE),
            Percentage = Death/(Alive + Death)) %>%
  knitr::kable(digits = 3)  

```

# Univariate Analysis
```{r}
# Distribution of Age
ggplot(mortality_data_EDA, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "black") +
  ggtitle("Age Distribution")

# Distribution of BMI (handling missing values)
ggplot(mortality_data_EDA %>% drop_na(bmi), aes(x = bmi)) +
  geom_histogram(binwidth = 1, fill = "darkgreen", color = "black") +
  ggtitle("BMI Distribution")

```

# Bivariate Analysis
```{r}

# Bivariate Analysis
ggplot(mortality_data_EDA, aes(x = gender, y = age, fill = outcome)) +
  geom_boxplot() +
  ggtitle("Age Distribution by Gender and Outcome")

```

**Age Distribution by Gender and Outcome:**
The boxplots show the age distribution for different genders and outcomes. We see that for both genders, the age distribution for patients who did not survive (outcome 1) is slightly higher. This could indicate that older age is a risk factor for mortality in this patient group.


# Exploring Outcomes
```{r}
# Age distribution by Outcome
ggplot(mortality_data_EDA, aes(x = age, fill = as.factor(outcome))) +
  geom_histogram(binwidth = 5, position = "dodge") +
  ggtitle("Age Distribution by Outcome")


# Blood Pressure by Outcome
ggplot(mortality_data_EDA, aes(x = systolic_blood_pressure, y = diastolic_blood_pressure, color = as.factor(outcome))) +
  geom_point(alpha = 0.5) +
  ggtitle("Blood Pressure by Outcome")

# BMI by Gender and Outcome
ggplot(mortality_data_EDA %>% drop_na(bmi), aes(x = bmi, fill = as.factor(outcome))) +
  geom_histogram(binwidth = 1, position = "dodge") +
  facet_wrap(~gender) +
  ggtitle("BMI by Gender and Outcome")

# Heart Rate by Outcome
ggplot(mortality_data_EDA, aes(x = heart_rate, fill = as.factor(outcome))) +
  geom_density(alpha = 0.7) +
  ggtitle("Heart Rate Distribution by Outcome")
```

**Age Distribution by Outcome:**

The histogram shows the age distribution for the two outcomes. We observe that the majority of patients are between 60 and 80 years old. There is a noticeable overlap in the age distribution of patients who survived and those who did not, suggesting that while age may be a factor, it's not the sole determinant of mortality.


**Blood Pressure by Outcome:**

This scatterplot shows systolic vs. diastolic blood pressure colored by outcome. The points are widely spread, indicating variability in blood pressure readings across the patient population. There doesn't seem to be a clear separation between the two outcomes based on blood pressure alone.


**BMI by Gender and Outcome:**

It shows a wide range of BMI values for both genders, with no clear pattern indicating that BMI is distinctly associated with the outcome. However, there are more data points for BMI for outcome 0, which may suggest a higher survival rate for patients with a lower BMI, although further statistical analysis is needed.


**Heart Rate Distribution by Outcome:**

The density plot for heart rate shows that the distribution of heart rates for patients who did not survive is slightly shifted towards the higher end compared to those who survived. This could suggest that a higher heart rate is associated with a higher risk of mortality.


# Comorbidities Analysis
```{r, fig.width = 14, fig.height = 6}
# Analyzing the presence of comorbidities by outcome
comorbidities <- c("hypertensive", "diabetes", "deficiencyanemias", "depression", "renal_failure", "copd", "hyperlipemia")

# Melt the data for easier plotting
mortality_long <- melt(mortality_data_EDA, id.vars = "outcome", measure.vars = comorbidities)

# Plotting comorbidities by outcome
ggplot(mortality_long, aes(x = variable, fill = as.factor(value))) +
  geom_bar(position = "fill") +
  facet_wrap(~outcome) +
  labs(x = "Comorbidity", y = "Count", fill = "Presence") +
  ggtitle("Distribution of Comorbidities by Outcome") +
  scale_y_continuous(labels = scales::percent)

```

**Distribution of Comorbidities by Outcome:**

The stacked bar plot shows the distribution of comorbidities by outcome. It appears that the proportion of patients with certain comorbidities like renal failure and COPD is higher among non-survivors. 


# Lab Results Analysis
```{r}
# Creatinine levels by outcome
ggplot(mortality_data_EDA %>% drop_na(creatinine), aes(x = creatinine, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  ggtitle("Creatinine Levels by Outcome")

# Lactic acid levels by outcome
ggplot(mortality_data_EDA %>% drop_na(lactic_acid), aes(x = lactic_acid, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  ggtitle("Lactic Acid Levels by Outcome")

# Urea nitrogen levels by outcome
ggplot(mortality_data_EDA %>% drop_na(urea_nitrogen), aes(x = urea_nitrogen, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  ggtitle("Blood Urea nitrogen Levels by Outcome")

# Leucocyte count by outcome
ggplot(mortality_data_EDA %>% drop_na(leucocyte), aes(x = leucocyte, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  ggtitle("Leucocyte Count by Outcome")

# Glucose levels by outcome
ggplot(mortality_data_EDA %>% drop_na(glucose), aes(x = glucose, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  ggtitle("Glucose Levels by Outcome")

ggplot(mortality_data_EDA %>% drop_na(anion_gap), aes(x = anion_gap, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  ggtitle("Anion Gap Levels by Outcome")

ggplot(mortality_data_EDA %>% drop_na(pco2), aes(x = pco2, fill = as.factor(outcome))) +
  geom_density(alpha = 0.5) +
  ggtitle("Pco2 Levels by Outcome")


```

**Blood Urea Nitrogen Levels by Outcome:**

Patients with higher levels seem to have a poorer outcome, as indicated by the longer tail in the distribution for non-survivors. High blood urea nitrogen levels can be indicative of renal insufficiency or failure, which is a known risk factor for mortality.

**Leucocyte Count by Outcome:**

The density plot shows a higher peak for leucocyte count among survivors (outcome 0) compared to non-survivors (outcome 1). However, there is a long tail in the distribution for non-survivors, suggesting that some patients who did not survive had very high leucocyte counts, which could indicate severe infection or systemic stress.

# Logistic Regression for Outcome Prediction
```{r}
# Logistic regression to assess the impact of various factors on outcome
# Select relevant variables for the logistic regression model
logistic_model <- glm(outcome ~ age + gender + bmi + hypertensive + diabetes + renal_failure + leucocyte +
                        urea_nitrogen + heart_rate, 
                      data = mortality_data, family = binomial())

# Summary of the model to check for significant variables
summary(logistic_model)

```