---
title: "Regeression"
output: 
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    code_folding: hide
---

```{r, message = FALSE}
library(tidyverse)
library(stargazer)
library(ROCR)
```

# About Regression

In this section, we want to generate a generalized linear model to see the relationship between patients' mortality status and predictive variables.

# Data Pre-processing

Firstly, to prepare the data for the regression, the data types of categorical variables will be transformed to factor. This subset of variables includes group (1 denoting the derivation group, 2 denoting the validation group), gender (1 denoting male, 2 denoting female), outcome (0 denoting alive, 1 denoting death), and the remaining comorbidity variables (0 denoting not having the disease, 1 denoting having the disease)

```{r, message = FALSE}
convert_to_factor <- function(df, columns) {
  df[columns] <- lapply(df[columns], factor)
  return(df)
}

mortality_cleaned = 
  read_csv("mortality_data_cleaned.csv") |>
  janitor::clean_names()  |>
  convert_to_factor(c("group", "gender", "outcome", "hypertensive", 
                      "atrialfibrillation", "chd_with_no_mi", "diabetes", 
                      "deficiencyanemias", "depression", "hyperlipemia", 
                      "renal_failure", "copd"))
```

# Generalized linear model

Due to the response variable (outcome) is binary variable, a generalized linear model (GLM) is chosen for regression analysis.

## Generalized Linear Model for Complete Dataset (Group1 and Group2)

Fit a logistic regression model using all patients data (including both the derivation and validation groups) and use the stepwise method to select predictors to be included in the model.

```{r, message = FALSE}
mortality_complete = 
  mortality_cleaned |>
  select(-group, -id)
```

```{r, message = FALSE}
glm_complete = glm(outcome ~ ., data = mortality_complete, family = binomial(link = logit))
```

```{r, results = "hide"}
stepwise_complete = step(glm_complete)
```

```{r}
stepwise_complete |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(
    caption = "Estimate and P-value of Generalized Linear Model Using All Patients Data", 
    col.names = c("Predictor", "Estimate", "P-value"),
    digits = 3
  )
```

The model fitted using data from all patients is:

$$
outcome \sim age+deficiencyanemias1+renal\_failure1+copd1\\
+heart\_rate+diastolic\_blood\_pressure+respiratory\_rate\\
+temperature+sp\_o2+urine\_output+mch+mchc+rdw+leucocyte\\
+platelets+lymphocyte+creatinine+urea\_nitrogen\\
+blood\_potassium+blood\_calcium+anion\_gap+magnesium\_ion\\
+bicarbonate+pco2
$$

It is worth noting that among the comorbidity variables, only deficiencyanemias, renal_failure, and copd are selected by model, and the estimates for all three are negative. 

## Generalized Linear Model for Derivation Dataset (Group1)

Fit a logistic regression model using patients data from group1 (derivation group) and use the stepwise method to select predictors to be included in the model.

```{r, message = FALSE}
mortality_derivation =
  mortality_cleaned |>
  filter(group == 1) |>
  select(-group, -id) 
```

```{r, message = FALSE}
glm_derivation = glm(outcome ~ ., data = mortality_derivation, family = binomial(link = logit))
```

```{r, results = "hide"}
stepwise_derivation = step(glm_derivation)
```

```{r}
stepwise_derivation |>
  broom::tidy() |>
  select(term, estimate, p.value) |>
  knitr::kable(
    caption = "Estimate and P-value of Generalized Linear Model Using Patients Data from Derivation Group", 
    col.names = c("Predictor", "Estimate", "P-value"),
    digits = 3
  )
```

The model fitted using data from derivation group is:

$$
outcome\sim age+gender2+hypertensive1+atrialfibrillation1+diabetes1\\ +deficiencyanemias1+hyperlipemia1+renal\_failure1+copd1+heart\_rate\\
+diastolic\_blood\_pressure+sp\_o2+mcv+rdw+platelets+lymphocyte\\
+creatinine+urea\_nitrogen+blood\_calcium+anion\_gap+lactic\_acid+pco2
$$

Among all the predictors selected, there are 7 comorbidity variables: hypertensive, atrialfibrillation, diabetes, deficiencyanemias, hyperlipemia, renal_failure and copd.

## Model Evaluation on Validation Group (Group2)

For the model generated using derivation group data, validate it on the validation group and plot Receiver Operating Characteristic (ROC) Curve.

```{r, massage = FALSE}
mortality_validation =
  mortality_cleaned |>
  filter(group == 2) |>
  select(-group, -id) 
```

```{r, massage = FALSE}
predict = predict(stepwise_derivation, newdata = mortality_validation, type = "response")
```

```{r, massage = FALSE}
pred = prediction(predictions = predict, labels = mortality_validation$outcome)
roc = performance(pred, "tpr", "fpr")
auc = performance(pred, measure = "auc")@y.values[[1]]
```

```{r}
plot(roc, colorize = F)
title(main = "The Receiver Operating Characteristic (ROC) Curve of Generalized Linear Model")
cat('AUC of the model on validation dataset:', auc)
```

According to the ROC plot and AUC value, the predictive performance of generalized linear model generated using group1 is satisfactory.