---
title: "Mortality Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: rows
    source_code: embed
runtime: shiny
---

```{r setup, include=FALSE}
library(flexdashboard)
library(tidyverse)
library(plotly)
library(shiny)
```


```{r,include=FALSE}

# Define a function to convert multiple columns to factors
convert_to_factor <- function(df, columns) {
  df[columns] <- lapply(df[columns], factor)
  return(df)
}

# Load the data and preprocess it
mortality_data <- read_csv("mortality_data.csv") %>%
  janitor::clean_names() %>%
  drop_na(outcome) %>%
  convert_to_factor(., c("group", "gendera", "outcome", "hypertensive", 
                         "atrialfibrillation", "chd_with_no_mi", "diabetes", 
                         "deficiencyanemias", "depression", "hyperlipemia", 
                         "renal_failure", "copd")) %>%
  rename(gender = gendera)  # Rename gendera to gender after conversion


# Imputing numerical columns with mean value 
numerical_columns <- sapply(mortality_data, is.numeric)
mortality_data[numerical_columns] <- lapply(mortality_data[numerical_columns], function(x) {
   ifelse(is.na(x), mean(x, na.rm = TRUE), x)
})

write_csv(mortality_data, "mortality_data_cleaned.csv")

mortality_data_EDA <- mortality_data %>%
  mutate(
    group = recode(group, `1` = "Group 1", `2` = "Group 2"),
    gender = recode(gender, `1` = "Male", `2` = "Female"),
    outcome = recode(outcome, `0` = "Alive", `1` = "Death")
  ) %>%
  mutate_if(is.factor, as.character)

  
```

```{r,include=FALSE}
# Change the factor variables for EDA 
# Manually recoding factors to their meaningful character values
mortality_data_EDA <- mortality_data %>%
  mutate(
    group = recode(group, `1` = "Group 1", `2` = "Group 2"),
    gender = recode(gender, `1` = "Male", `2` = "Female"),
    outcome = recode(outcome, `0` = "Alive", `1` = "Death")
  ) %>%
  mutate_if(is.factor, as.character) # Converts all remaining factors to characters

# Function to recode factor variables to 'Yes' or 'No'
convert_factors_to_yes_no <- function(df, comorbidity_columns) {
  df <- df %>%
    mutate(across(all_of(comorbidity_columns), ~ ifelse(as.character(.) == "1", "Yes", "No")))
  return(df)
}

# List of comorbidity columns to convert
comorbidity_columns <- c("hypertensive", "atrialfibrillation", "chd_with_no_mi", 
                         "diabetes", "deficiencyanemias", "depression", 
                         "hyperlipemia", "renal_failure", "copd")

# Apply the function to the mortality_data_EDA dataframe
mortality_data_EDA <- convert_factors_to_yes_no(mortality_data_EDA, comorbidity_columns)

```


Demographic Characteristics
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------


```{r}
selectInput(
    inputId = "char_index",
    label = h3("Select Index"),
    choices = c("age", "gender", "bmi"),
    selected = "age"
  )
```




Row{data-height=650}
-----------------------------------------------------------------------

### Demographic Characteristics

```{r}
renderPlotly({

variable <- input$char_index
    plot_ly(mortality_data_EDA %>% drop_na(.data[[variable]]), 
            x = ~.data[[variable]], color = ~as.factor(outcome),
            type = "scatter", mode = "lines", fill = "tozeroy") %>%
      layout(title = paste(variable, "Levels by Outcome"),
             xaxis = list(title = variable),
             yaxis = list(title = "Density"))
  
})

```

Row {.tabset .tabset-fade}
-------------------------------------


### Description


This is the world map showing the life expectancy (years) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * North America, Western and Northern Europe, Japan, New Zealand and Australia show longer life expectancy compared with other regions.
  
\ \par
  * Central and Southern Africa have relatively shorter life expectancy because the color is lighter,
  
### Boxplot

```{r}
renderPlotly({
  variable <- input$char_index
  
  plot_ly(mortality_data_EDA %>% drop_na(.data[[variable]]), 
          x = ~as.factor(outcome),
          y = ~.data[[variable]],
          type = "box") %>%
    layout(title = paste(variable, "Levels by Outcome"),
           xaxis = list(title = "Outcome"),
           yaxis = list(title = variable))
})

```


Vital Signs
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------


```{r}
selectInput(
    inputId = "vital_sign_index_x",
    label = h3("Select Index"),
    choices = c("heart_rate", "systolic_blood_pressure", "diastolic_blood_pressure", "respiratory_rate", "temperature", "sp_o2", "urine_output"),
    selected = "systolic_blood_pressure"
  )


```

```{r}
selectInput(
    inputId = "vital_sign_index_y",
    label = h3("Select Index"),
    choices = c("heart_rate", "systolic_blood_pressure", "diastolic_blood_pressure", "respiratory_rate", "temperature", "sp_o2", "urine_output"),
    selected = "diastolic_blood_pressure"
  )


```


Row{data-height=650}
-----------------------------------------------------------------------

### Vital Signs

```{r}
renderPlotly({

  vital_sign_index_x <- input$vital_sign_index_x
  vital_sign_index_y <- input$vital_sign_index_y

  
  plot_ly(mortality_data_EDA %>% drop_na(.data[[vital_sign_index_x, vital_sign_index_y]]), 
          x = ~.data[[vital_sign_index_x]], 
          y = ~.data[[vital_sign_index_y]], 
          color = ~as.factor(outcome),
          type = "scatter",
          mode = "markers",
          marker = list(size = 8, opacity = 0.5)) %>%
    layout(title = paste(vital_sign_index_x, "vs", vital_sign_index_y, "by Outcome"),
           xaxis = list(title = vital_sign_index_x),
           yaxis = list(title = vital_sign_index_y))
  
})

```



Row {.tabset .tabset-fade}
-------------------------------------


### Description


This is the world map showing the life expectancy (years) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * North America, Western and Northern Europe, Japan, New Zealand and Australia show longer life expectancy compared with other regions.
  
\ \par
  * Central and Southern Africa have relatively shorter life expectancy because the color is lighter,
  
### Boxplot

```{r}
renderPlotly({
  variable <- input$vital_sign_index
  
  plot_ly(mortality_data_EDA %>% drop_na(.data[[variable]]), 
          x = ~as.factor(outcome),
          y = ~.data[[variable]],
          type = "box") %>%
    layout(title = paste(variable, "Levels by Outcome"),
           xaxis = list(title = "Outcome"),
           yaxis = list(title = variable))
})

```

Comorbidities
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------


```{r}
selectInput(
    inputId = "como_index_x",
    label = h3("Select Index"),
    choices = c("hypertensive", "atrialfibrillation", "chd_with_no_mi", "diabetes", "depression", "hyperlipemia", "hyperlipemia", "copd"),
    selected = "diabetes"
  )
```

```{r}
selectInput(
    inputId = "como_index_y",
    label = h3("Select Index"),
    choices = c("hypertensive", "atrialfibrillation", "chd_with_no_mi", "diabetes", "depression", "hyperlipemia", "hyperlipemia", "copd"),
    selected = "depression"
  )


```


Row{data-height=650}
-----------------------------------------------------------------------

### Comorbidities

```{r}
renderPlotly({
    como_index_x <- input$como_index_x
  como_index_y <- input$como_index_y
  
  plot_ly() %>%
    add_histogram(data = mortality_data_EDA, x = ~.data[[como_index_x]], 
                  color = ~as.factor(outcome), 
                  opacity = 0.5, 
                  name = "Outcome 1") %>%
    add_histogram(data = mortality_data_EDA, x = ~.data[[como_index_y]], 
                  color = ~as.factor(outcome), 
                  opacity = 0.5, 
                  name = "Outcome 2") %>%
    layout(barmode = "overlay",
           title = paste("Histograms of", como_index_x, "and", como_index_y, "by Outcome"),
           xaxis = list(title = "Values"),
           yaxis = list(title = "Frequency"))
})

```

Row {.tabset .tabset-fade}
-------------------------------------


### Description


This is the world map showing the life expectancy (years) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * North America, Western and Northern Europe, Japan, New Zealand and Australia show longer life expectancy compared with other regions.
  
\ \par
  * Central and Southern Africa have relatively shorter life expectancy because the color is lighter,
  
### Boxplot

```{r}
renderPlotly({
  como_index_x <- input$como_index_x
 como_index_y <- input$como_index_y
  
  plot_ly() %>%
    add_boxplot(data = mortality_data_EDA, x = ~as.factor(outcome), y = ~.data[[como_index_x]], 
                name = como_index_x, boxpoints = "all", jitter = 0.3, pointpos = -1.8) %>%
    add_boxplot(data = mortality_data_EDA, x = ~as.factor(outcome), y = ~.data[[como_index_y]], 
                name = como_index_y, boxpoints = "all", jitter = 0.3, pointpos = -0.6) %>%
    layout(title = paste("Box Plots of", como_index_x, "and", como_index_y, "by Outcome"),
           xaxis = list(title = "Outcome"),
           yaxis = list(title = "Values"))
})

```

Lab Results Analysis
===================================== 

Column {.sidebar}
-----------------------------------------------------------------------


```{r}
selectInput(
    inputId = "index",
    label = h3("Select Index"),
    choices = c("creatinine", "lactic_acid", "urea_nitrogen", "leucocyte", "glucose", "anion_gap", "pco2"),
    selected = "creatinine"
  )


```




Row{data-height=650}
-----------------------------------------------------------------------

### Lab Results Analysis

```{r}
renderPlotly({

variable <- input$index
    plot_ly(mortality_data_EDA %>% drop_na(all_of(variable)), 
            x = ~.data[[variable]], color = ~as.factor(outcome),
            type = "scatter", mode = "lines", fill = "tozeroy") %>%
      layout(title = paste(variable, "Levels by Outcome"),
             xaxis = list(title = variable),
             yaxis = list(title = "Density"))
  
})

```

Row {.tabset .tabset-fade}
-------------------------------------


### Description


This is the world map showing the life expectancy (years) of each country from 2000 to 2016. 

\ \par
Take the year of 2015 as an example,

\ \par
  * North America, Western and Northern Europe, Japan, New Zealand and Australia show longer life expectancy compared with other regions.
  
\ \par
  * Central and Southern Africa have relatively shorter life expectancy because the color is lighter,
  
### Boxplot

```{r}
renderPlotly({
  variable <- input$index
  
  plot_ly(mortality_data_EDA %>% drop_na(.data[[variable]]), 
          x = ~as.factor(outcome),
          y = ~.data[[variable]],
          type = "box") %>%
    layout(title = paste(variable, "Levels by Outcome"),
           xaxis = list(title = "Outcome"),
           yaxis = list(title = variable))
})

```
